P4IRS: An Intermediate Representation and Compiler for Parallel and Performance-Portable Particle Simulations

### 2024-01-08 manuscript.tex
Improve results section
Signed-off-by: Rafael Ravedutti <rafael.r.ravedutti@fau.de>

The icc (ICC) 2021.10.0 20230609 compiler was used for generating the CPU code, with the \emph{-Ofast -xCORE-AVX512 -qopt-zmm-usage=high} flags for Ice Lake (AVX512) and \emph{-Ofast -march=core-avx2} flags for Milan (AVX2).
For CUDA, we use the NVCC compiler v12.1.r12.1/compiler.32688072, with same flags as the ones used for the Milan system.
We make use of the Likwid toolkit V5.3 \cite{likwid} to measure HPM counters in the target CPU (Ice Lake) for the single core experiments, where P4IRS is able to properly generate the instrumentation markers and MD-Bench kernels are already instrumented.

Experiments for \ac{MD} use the standard Copper FCC lattice (see \autoref{fig:copper_fcc_lattice}) testcase from miniMD/MD-Bench, and we use our results from MD-Bench as a performance reference since it is a well optimized code for \ac{MD}.
The \ac{DEM} test case is a "Settling Spheres" benchmark (see \autoref{fig:settling_spheres}), on which several spheres fall into a "ground" (half-space), thus generating a "bed" of spherical particles.
Due to its multiple shapes and dynamics that consists of many interactions among spheres and half-spaces, it should provide a good reference for DEM simulations.
We compare our single-core results from P4IRS with MESA-PD using 2 MPI ranks since it is not possible to run this setup in MESA-PD using a single rank.
GPU results are not available for MESA-PD as it is not capable of generating GPU code, but we show GPU performance results for P4IRS.

We also show weak-scaling results to demostrate P4IRS scaling capabilities.
For this we use regular domain-partitioning since these experiments do not require sophisticated load-balancing algorithms (Copper lattice is an homogeneous testcase in all dimensions, and Settling Spheres can obtain good weak-scaling by regularly partitioning the domain in X and Y dimensions).
We use double floating-point precision and display results when enabling and disabling Newton's 3rd Law optimization (half neighbor-lists).
For \ac{MD} we show results for the Verlet Lists case since it is widely employed in such simulations, and for \ac{DEM} we stick to Linked Cells as this is the algorithm used in MESA-PD.

@@

\autoref{fig:md_single_core_runtimes} display the runtime measurements for P4IRS and MD-Bench simulations of the Copper FCC lattice case with $32^{3}$ unit cells (4 atoms per unit-cell).
It is noticeable that the kernels generated by P4IRS achieve competitive results to MD-Bench, a hand-tuned implementation for MD cases, which demonstrates P4IRS capabilities to generate fast code.
For Ice Lake, performance is slightly better for MD-Bench using \ac{FN}, but P4IRS provide a slightly faster result when using \ac{HN}.
P4IRS code tends to be build neighbor-lists faster but the force computation is slower, but runtime differences vary in the range of 10\% for the measured runtime, which is still not too significant and can vary when using different compilers and flags.
In Milan, P4IRS is faster than MD-Bench for the \ac{FN} case, and results are pretty similar when using \ac{HN} case.
The force calculation is still slower for P4IRS with \ac{HN} when compared to MD-Bench, but this is compensated by the less time spent building the neighbor lists.

@@

\autoref{tab:md_single_core_stats} displays HPM measurements for the force calculation kernels in Ice Lake with AVX512.
The important thing to notice is that the compiler was not able to vectorize the cases with \ac{HN}, but the runtimes are still smaller for these cases.
The performance with \ac{HN} is significantly worse but it is compensated by the fact it computes roughly half the interactions.
The arithmetic instructions ratio for all cases lie between 50-55\%, which means that about half of the instructions are used for organizing data into the vector registers.
For more computational-intensive kernels, the arithmetic ratio should increase and the measured AVX512/Scalar performance should have a more significant impact in the runtime.

@@

\autoref{fig:md_single_gpu_runtimes} displays the runtime measurements for P4IRS and MD-Bench simulations in the A40 and A100 GPUs from the Alex cluster.
MD-Bench still do not have a GPU version with \ac{HN}, so we just display it using \ac{FN}.
For A40, there is no significant difference for all runs, with MD-Bench being slightly faster.
The performance difference is larger for A100, where MD-Bench is significantly faster.
The usage of atomics indeed harms the performance since \ac{HN} computes half the interactions in the same amount of time in A40, and this impact is significant worse in A100 since it takes about twice the amount of time to compute half the interactions.
Note again that this is strictly dependent on the arithmetic intensity of the kernel, which is relatively small for the used \ac{LJ} kernels, for other cases it may be worth to make use of \ac{HN}.

@@

\autoref{fig:md_weak_scaling_fritz} shows the results for P4IRS weak-scaling in the Fritz supercomputer.
For each node, a system of $128^3$ unit cells is included, with roughly 8.388.608 atoms per node (or 116.508 atoms per CPU core).
Results demonstrate that P4IRS achieve near perfect weak-scaling in all 64 nodes of the Fritz cluster (4608 CPU cores), which suggests it is fully capable of generating scalable applications to run on modern supercomputers.

>>>>

The icc (ICC) 2021.10.0 20230609 compiler was used for CPU experiments (except in the JUWELS cluster), with the \emph{-Ofast -xCORE-AVX512 -qopt-zmm-usage=high} flags for Fritz / Ice Lake (AVX512) and \emph{-Ofast -march=core-avx2} flags for Milan (AVX2).
For CUDA, we use the NVCC compiler V12.1.r12.1 for the single GPU experiments, and the same flags used for Milan (AVX2) since it is the host processor in the systems with the A40/A100 GPUs.
In the JUWELS Booster, we used the Intel(R) oneAPI DPC++/C++ Compiler 2023.2.0 (2023.2.0.20230721) together with NVCC V12.2.91, and the same flags as used for Milan since both CPUs have AVX2 with FMA as the largest SIMD vector extension available.
We make use of the Likwid toolkit V5.3 \cite{likwid} to measure HPM counters in the target CPU (Ice Lake) for the single core experiments, where P4IRS is able to properly generate the instrumentation markers and MD-Bench kernels are already instrumented.

Experiments for \ac{MD} use the standard Copper FCC lattice (see \autoref{fig:copper_fcc_lattice}) testcase from miniMD/MD-Bench, and we use our results from MD-Bench as a performance reference since it is a well optimized code for \ac{MD}.
The \ac{DEM} test case is a "Settling Spheres" benchmark (see \autoref{fig:settling_spheres}), on which several spheres fall into a "ground" (half-space), thus generating a "bed" of spherical particles.
Due to its multiple shapes and dynamics that consists of many interactions among spheres and half-spaces, it should provide a good reference for DEM simulations.
We compare our single-core results from P4IRS with MESA-PD using 2 MPI ranks since it is not possible to run this setup in MESA-PD using a single rank.
GPU results are not available for MESA-PD as it is not capable of generating GPU code, but we show GPU performance results for P4IRS comparing different code generation strategies.

We also show weak-scaling results to demostrate P4IRS scaling capabilities.
For this we use regular domain-partitioning since these experiments do not require sophisticated load-balancing algorithms (Copper lattice is an homogeneous testcase in all dimensions, and Settling Spheres can obtain acceptable weak-scaling by regularly partitioning the domain in X and Y dimensions).
We use double floating-point precision and display results when enabling and disabling Newton's 3rd Law optimization (half neighbor-lists).
For \ac{MD} we show results for the Verlet Lists case since it is widely employed in such simulations, and for \ac{DEM} we stick to Linked Cells as this is the algorithm used in MESA-PD.

@@

\autoref{fig:md_single_core_runtimes} display the runtime measurements for P4IRS and MD-Bench simulations of the Copper FCC lattice case with $32^{3}$ unit cells (4 atoms per unit-cell).
It is noticeable that the kernels generated by P4IRS achieve competitive results in comparison to MD-Bench, a hand-tuned implementation for MD cases, thus demonstrating P4IRS capabilities to generate fast code.
For Ice Lake, performance is slightly better for MD-Bench using \ac{FN}, with about 10\% speedup for the force calculation but 25\% worse performance for the neighbor-lists creation.
When using \ac{HN}, the force calculation time is roughly the same for both versions, but P4IRS still keeps the advantage when bulding the neighbor-lists, resulting in a 4\% better performance for P4IRS in the overall runtime.
Note that the code from P4IRS tends to be build neighbor-lists faster for these experiments but the force computation is slower for \ac{FN}, and the runtime differences vary in the range of 10\% for the measured runtime, which is not too significant and can vary when using different compilers and flags.

In Milan, P4IRS is faster than MD-Bench for the \ac{FN} case, and the advantage also comes from the neighbor-lists creation.
For \ac{HN}, the results are pretty similar, but it is possible to see that P4IRS still builds the neighbor-lists slightly faster, where MD-Bench has a moderately better performance for the force computation.
Overall, we can assume that P4IRS is able to achieve good base performance when running in a single CPU core, while providing a more flexible and high-level description of the problem which is not the case for the hand-tuned and hard-coded C implementation from MD-Bench.

@@

For a deeper evaluation of both versions for the \ac{MD} case, \autoref{tab:md_single_core_stats} displays HPM measurements for the force calculation kernels in Ice Lake with AVX512.
The important thing to notice is that the compiler was not able to vectorize the cases with \ac{HN}, but the runtimes are still smaller for these cases.
The arithmetic instructions ratio for all cases lie between 50-55\%, which means that about half of the instructions are used for organizing data into the vector registers.
For more computational-intensive kernels, the arithmetic ratio should increase and the measured AVX512/Scalar performance should have a more significant impact in the runtime.

The performance with \ac{HN} is significantly worse but it is compensated by the fact it computes roughly half the interactions.
Another important thing to consider is the introduction of the Intel patch to mitigate the Downfall vulnerability (also known as Gather Data Sampling), which significantly affects the performance of the gather instructions in the Intel CPUs and therefore affect the \ac{FN} cases performance with AVX512 vectorization.
In particular, the separation of concerns introduced by P4IRS together with code generation techniques can assist on dealing with such issues by avoiding gather instructions, which can be done by producing explicit SIMD intrinsics to gather the data with separate instructions to load and organize (through permutatation, shuffles and swizzles) data into the vector registers, or using different algorithms such as the Cluster Pair from GROMACS, which keeps atoms data in an array-of-struct-of-arrays (AoSoA) fashion, thus requiring only simple load instructions to build-up the vector registers.

@@

\autoref{fig:md_single_gpu_runtimes} displays the runtime measurements for P4IRS and MD-Bench simulations in the A40 and A100 GPUs.
MD-Bench still do not have a GPU version with \ac{HN}, so we just display it using \ac{FN}.
For A40, MD-Bench shows a slightly better performance (about 3\%) for computing the forces while P4IRS with \ac{FN} also provides slighly better performance (about 7\%) for building the neighbor-lists.
In the overall performance, MD-Bench is faster only by about 2\% in comparison to P4IRS for the \ac{FN} case, which is not too significant and corroborates our claim that P4IRS is able to generate efficient GPU code.
For the \ac{HN} case, the performance is not even close to a factor of two as one would expect, which can be attributed to the usage of atomics, and this can be different for more arithmetic-intensive kernels (the \ac{LJ} kernel used has significantly less arithmetic instructions compared to realistic cases) since the overall amount of operations reduced should be larger.
For A100, however, the performance difference is larger, where MD-Bench performance is about 14\% faster in comparison to P4IRS.
When using the \ac{HN}, however, we achieve very close performance to MD-Bench with \ac{FN}, however it is noticeable that the speedup is very far from the factor of two as well.
Despite the observed performance divergence in A100, it is possible to see that P4IRS code still has good performance in comparison to our hand-tuned CUDA implementation, supporting our claim in generating performance-portable kernels.

@@

\autoref{fig:md_weak_scaling_fritz} shows the results for P4IRS weak-scaling in the Fritz supercomputer.
For each node, a system of $128^3$ unit cells is included, with roughly 8.388.608 atoms per node (or 116.508 atoms per CPU core).
Since the \emph{multinode} partition in Fritz has the maximum of 64 nodes (which results in 4608 CPU cores, no hyper-threading), this is the maximum amount we scale up our problem to.
It is possible to see that we achieve perfect weak-scaling in all the settings we use, and the \ac{HN} version is significantly faster than the \ac{FN} version.
Notice that atomic operations are not required because we only make use of parallelism through spatial decomposition and message communication with MPI, thus we use distributed-memory parallelism strategy even within the same node and there is no possibility of race conditions.
Based on our results, it is clear that P4IRS is able to generate scalable code on modern CPU clusters, but problems that are not too homogeneous may need proper mechanisms for balancing the workload among different MPI ranks.
